---
title: "Mixed Model Lab 7"
output:
  html_document:
    df_print: paged
---
```{r}
# install.packages("AmesHousing")
install.packages("rsq")
```


```{r setup, include = FALSE}
knitr::opts_chunk$set(
  comment = "",
  results = "hold",
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 5,
  fig.align = "center"
)
```
Read in packages
```{r, message = FALSE, warning=FALSE}
library(rio)
library(dplyr)
library(tidyr)
library(magrittr)
library(ggplot2)
library(pROC)
library(AmesHousing)
library(tidyverse)
library(rcompanion)
library(lme4)
library(rsq)
library(janitor)
library(here)
```

```{r}
autot <- clean_names(read_csv(here("data", "Automobile_data.csv"),
  na = c("?", "", "NA", "N/A", "Na"),
))
autot <- autot %>%
  mutate_at(
    vars(fuel_type, body_style, make),
    list(factor)
  )
```

## Comparing fixed factors versus random effects

1.  Build a model of curb_weight as a function of the factor "make".

```{r}
mmod1 <- lmer(curb_weight ~ (1 | Maked), data = autot)
mmod1

```



2. Create a vector of the coefficients from the model above but start the vector with the coefficient (0) for the base level Alfa-Romeo (i.e., the first element of the vector should be zero, replacing the model intercept). Calculate the sd of this vector.

3. Build a model of curb_weight as a function of the random effect "make" and comment on the findings.

4. Extract the estimate of the effect of each make on curb_weight as a data frame.

5. Compare the sd of coefficients from the fixed effect (question 2) with the sd of the random effect and comment on the difference.

6. Visualise the estimates for both models by using a histogram of Make coefficients.

7. Now try creating a model of curb_weight as a function of the random effect make and the fixed effects length and horsepower (remember to drop NAs).

8. Create a confidence interval for the estimates in the model and comment on them.

9. Compare the impact of fixed with random effects, in particular we need to compare minimum (and predicted) effect associated to 1 sd change in predictor.

  a. Find the sd for length multiply this by the lower estimate from the CI and then divide this by the mean curb_weight and then comment on the obtained decimal (convert to a percentage) - this will give the minimum effect associated with 1 sd change in the predictor

  b. repeat (a.) for horsepower and the random effect. For random effect we just need to use the lower estimate for sig01 divided by mean curb_weight (as sig01 represents the standard deviation of the random effect).
  
10. Find the R-squared.

11. Use the R-squared to calculate the adjusted R-squared

Recall adjusted R-squared accounts for including more variables in a model.  This is given by the formula:

$$
1-(1-R^2)(n-1/n-k)
$$

where n=number of cases, k=number of parameters in the model.

12. Compare these values to a model does not include make but does include horsepower and length.