---
title: "Lab 5" 
output:
  html_document:

    highlight: textmate
    toc_depth: 2
---
```{r}
# install.packages("investr")
# install.packages("rio")
# install.packages("rcompanion")
# install.packages("ggcorrplot")
# install.packages("corrr")
# install.packages("effects")
install.packages("magrittr")
```


```{r setup, include = FALSE}
knitr::opts_chunk$set(
  comment = "",
  results = "hold",
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 5,
  fig.align = "center"
)
```


### Set up from previous labs
```{r, message = FALSE,comment=FALSE, warning=FALSE}
library("MASS")
library("sandwich")
library("investr")
library("rio")
library("pROC")
library("nnet")
library("tidyverse")
library("magrittr")
library("here")
library("janitor")
library("readxl")
library("rcompanion")
library("ggcorrplot")
library("corrr")
library("effects")
```


```{r}
# 
trees2 <- trees %>%
  mutate(Diameter = Girth) %>%
  dplyr::select(Height, Diameter, Volume)

autot <- clean_names(read_csv(here("data", "Automobile_data.csv"),
  na = c("?", "", "NA", "N/A", "Na"),
))

comic_data <- read_csv(here("data", "comic_characters.csv"))
comic_data <- comic_data %>%
  mutate_at(
    vars(ID, ALIGN, EYE, HAIR, ALIVE, MONTH_FA, UNIVERSE),
    list(factor)
  )
clean_comic_data <- comic_data %>%
  drop_na()
comic_model_data <- clean_comic_data %>%
  filter(YEAR_FA >= 1956 & YEAR_FA <= 1999)
comic_model_data$ALIGN <- droplevels(comic_model_data$ALIGN)
```

### Question 1
**Exercise (a):** Find the confidence intervals for the coefficients in linmod4 and linmod5 (given below) and interpret these in terms of the dataset.

```{r}
linmod4 <- lm(highway_mpg ~ width + curb_weight + fuel_type + body_style, data = autot)

linmod5 <- glm(as.factor(aspiration) ~ drive_wheels + engine_size + horsepower, family = "binomial", data = autot)

confint(linmod4)
summary(linmod4)
confint(linmod5)
summary(linmod5)
```


### Question 2
**Exercise (b)** calculate the robust confidence intervals for linmod3 (below) and interpret in terms of the dataset.

```{r}

trees2 <- trees2 %>%
  mutate(logheight = log(Height)) %>%
  mutate(logvol = log(Volume)) %>%
  mutate(logdiam = log(Diameter))
linmod3 <- lm(logvol ~ logheight + logdiam, data = trees2)

# Calculate the robust confidence intervals for linmod3, and interpret in terms of the dataset

std_err <- sqrt(diag(vcovHC(linmod3, type = "HC0")))
r_est <- cbind(
 Estimate = coef(linmod3), "Robust SE" = std_err,
 "Pr(>|z|)" = 2 * pnorm(abs(coef(linmod3) / std_err), lower.tail = FALSE),
 LL = coef(linmod3) - 1.96 * std_err,
 UL = coef(linmod3) + 1.96 * std_err
)
r_est

summary(trees2)
summary(linmod3)
```


### Question 3
For comic_linmod3 below, estimate the confidence interval for the probability that the UNIVERSE is MARVEL. 

```{r}
comic_linmod3 <- glm(UNIVERSE ~ YEAR_FA, family = "binomial", data = comic_model_data)

confint(comic_linmod3)

summary(comic_linmod3)
```

### Question 4
For multi_comic model below, use the effects function to visualise confidence intervals for Align based on year_fa. How the range of predicted probabilities for each class changes as year_fa changes.

```{r}
multi_comic <- multinom(ALIGN ~ ALIVE + YEAR_FA, data = comic_model_data)

summary(multi_comic)
```

### Question 5
Calculate the confidence interval for the mean number of appearances of good characters that are alive and introduced in 1990 using Poisson and QuasiPoission models below. Compare the two intervals.
```{r}
comic_poissonmod <- glm(APPEARANCES ~ YEAR_FA + ALIGN + ALIVE, data = comic_model_data, family = "poisson")
comic_quassipoissonmod <- glm(APPEARANCES ~ YEAR_FA + ALIGN + ALIVE, data = comic_model_data, family = "quasipoisson")

summary(comic_poissonmod)
summary(comic_quassipoissonmod)
```
